---
title: "WA_DEMs 10m Stack - Derive and store terrain metrics"
description: "At 10m -  generates terrain layers for Washington State"
author: "Aji John"
format: html
editor: visual
---

## Overview

This notebook provides a detailed guide for processing 10-meter resolution Digital Elevation Models (DEMs) for Washington State (WA) using R. Includes downloading, cropping, and generating terrain metrics such as slope, roughness, and Topographic Wetness Index (TWI) for HUC8 watersheds.

\#\@\@ Objectives

Clip and process DEM data for HUC8 regions in WA. Calculate terrain attributes including slope, roughness, TWI, and curvature. Automate the processing of all HUC8 watersheds in the state.


## Running Code

```{r}
library(terra)
library(tidyterra)
library(stringr)
library(dplyr)
library(randomForest)
library(caret)
library(ggplot2)
library(lme4)
set.seed(11)
library(elevatr)  
library(nhdplusTools)
library(MultiscaleDTM)
library(whitebox)

#terraOptions(chunksize = 1e6, memfrac = 0.9)  # Adjust memory fraction and chunk size
```
## Load Preprocessed data - primarily utility functions

```{r}
load('./fun.RData')
```

## Load and Prepare Shapefiles

```{r}
wa_shapes <- vect("data/wa_shp_diss.gpkg")
wa_huc8s <- vect("data/Hydrography/WA_HUC8.gpkg")
wa_huc8s$name
```


## Data Processing Workflow

-   Prepare Boundary Data
-   Obtain HUC8 Boundaries

```{r}
library(sf)
library(tigris) 
# Get Washington state boundary
wa_boundary <- states(cb = TRUE) %>% 
  filter(STUSPS == "WA") %>% 
  st_transform(crs = 4326)  # Ensure it's in WGS84 CRS

# Get all HUC8 boundaries that intersect Washington State
huc8_regions <- get_huc(AOI = wa_boundary, type = "huc08")
```

## Define File Paths

Ajust as needed

```{r}
datapath <- paste0(getwd(), "/data/WA_10m_R1")
```

## Define DEM Download Function - clip, calculate and save

terr_stackXXX XXXX is the name of the basin

Key tasks:

-DEM Cropping and Masking: Crops DEM to the watershed boundary. -Terrain Metric Calculations: Slope, roughness, curvature, TWI. -WhiteboxTools Integration: Filling depressions, flow accumulation. -Saving Outputs: Exports combined rasters as GeoTIFF files.

```{r}
calculate_tpi <- function(dem, scale) {
  # Create a circular matrix for the specified scale
  matrix_radius <- focalMat(dem, d = scale, type = "circle")
  
  # Calculate mean elevation within the defined scale
  mean_elevation <- focal(dem, w = matrix_radius, fun = mean, na.rm = TRUE)
  
  # Calculate TPI as the difference between cell elevation and mean neighborhood elevation
  tpi <- dem - mean_elevation
  return(tpi)
}
```

```{r}


dem <- rast("./data/WA_10mDEM_3740.tif")

# Function to download DEM for a single HUC8 region
download_huc8_dem_name <- function(huc8_boundary, huc8_id,huc8_name, zoom_level = 11) {
  tryCatch({
    
    #dem <- get_elev_raster(locations = huc8_boundary, z = 11, clip = "locations")
    # Crop the DEM to the extent of the HUC8 boundary
    # Define the target CRS (UTM Zone 10N, WGS 84)
    target_crs <- "EPSG:3740"

    # Reproject the polygon to the target CRS
     huc8_boundary <- st_transform(huc8_boundary, crs = 3740)

    dem_cropped <- crop(dem, huc8_boundary)
    # Step 4: Mask the DEM to the exact shape of the HUC8 boundary
    dem_cropped <- mask(dem_cropped, huc8_boundary)

    # Step 2: Calculate slope in radians
    slope <- terrain(dem_cropped,  "slope", unit = "radians")
    roughness <- terrain(dem_cropped,  "roughness")
    
    # Calculate slope in degrees
    #slope_degrees <- terrain(dem_cropped, v = "slope", unit = "degrees")

    # Calculate slope in radians (useful for certain hydrological calculations, e.g., TWI)
    slope_radians <- terrain(dem_cropped, v = "slope", unit = "radians")

    writeRaster(slope_radians, "./working_slope.tif",  gdal=c("COMPRESS=LZW"),overwrite=TRUE)
   
    tpi <- terrain(dem_cropped,  "TPI")
    vrm <- VRM(dem_cropped, w = c(3, 3))
  
    # Calculate the specific catchment area (SCA)
    #flowdir <- terrain(dem_cropped, "flowdir")
    #weight<- cellSize(dem_cropped,transform=FALSE,unit="m")

    #The algorithm is an adaptation of the one proposed by Zhou at al, 2019.
    # Calculate flow accumulation (which is equivalent to Specific Catchment Area)
    #flow_accumulation <- flowdir * weight
    
    file_name ='workingdem.tif'
    writeRaster(dem_cropped, file_name,  gdal=c("COMPRESS=LZW"),overwrite=TRUE)
    # Step 2: Fill depressions in the DEM using WhiteboxTools
    wbt_fill_depressions("./workingdem.tif", "./filled_dem.tif")
    
    filledeelevation <- rast("./filled_dem.tif")

    # Step 3: Calculate D8 flow accumulation
    wbt_d8_flow_accumulation("./filled_dem.tif", "./flow_accumulation.tif", out_type = "cells")
    wbt_d_inf_flow_accumulation("./filled_dem.tif", "./flow_accumulationdinf.tif")
    
    cat("d8 and Inf flow  ",  "\n")
    
    slp <- rast("./working_slope.tif")
    # Load the flow accumulation (SCA) raster
    sca <- rast("./flow_accumulation.tif")
    slp <- rast("./working_slope.tif")
    # Calculate TWI
    # Step 4: Calculate TWI using the SCA and slope
    twi <- log(sca / tan(slp))
    twi[is.infinite(twi)] <-  NA
    
    cat("TWI  ",  "\n")

    # Calculate profile curvature with a specified window size
    # Calculate profile curvature with a specified window size
    profile_curv <- Qfit(dem_cropped, metrics = c("profc"), w = c(3, 3))
    cat("Prof curv @3  ",  "\n")

    # Calculate profile curvature with a specified window size
    profile_curv9by9 <- Qfit(dem_cropped, metrics = c("profc"), w = c(9, 9))
    cat("Prof curv @9  ",  "\n")
    profile_curv17by17 <- Qfit(dem_cropped, metrics = c("profc"), w = c(17, 17))
    cat("Prof curv @17  ",  "\n")
   # profile_curv100by100 <- Qfit(dem_cropped, metrics = c("profc"), w = c(99, 99))
    #cat("Prof curv @99 ",  "\n")
    
    plan_curv <- Qfit(dem_cropped, metrics = c("planc"), w = c(3, 3))
    #plot(dem)
    cat("Plan curv @3 ",  "\n")
    plan_curv9by9 <- Qfit(dem_cropped, metrics = c("planc"), w = c(9, 9))
     cat("Plan curv @3 ",  "\n")
    plan_curv17by17 <- Qfit(dem_cropped, metrics = c("planc"), w = c(17, 17))
     cat("Plan curv @17 ",  "\n")
    #plan_curv49by49 <- Qfit(dem_cropped, metrics = c("planc"), w = c(49, 49))
    #cat("Plan curv @49 ",  "\n")
    
    # Calculate TPI at different scales
    tpi_100m <- calculate_tpi(dem_cropped, scale = 100)
     cat("TPI @100 ",  "\n")
    tpi_200m <- calculate_tpi(dem_cropped, scale = 200)
     cat("TPI @200 ",  "\n")
   # tpi_500m <- calculate_tpi(dem_cropped, scale = 500)
  #   cat("TPI @500 ",  "\n")

    
    # Define output stream raster file path
    streams_output <- "./streams.tif"

    # Set a threshold for stream extraction (e.g., 1000). Adjust as needed.
    threshold <- 1000

    # Run Extract Streams
   wbt_extract_streams("./flow_accumulationdinf.tif", output = streams_output, threshold = threshold)
    cat("Ext streams",  "\n")
   streams <- rast("streams.tif")

     # DTW
    # Mask the DEM using the stream layer to keep elevations only at stream cells
    stream_elevation <- mask(filledeelevation, streams)

     cat("Stream elevation",  "\n")
    # Use `terra::focal` with a larger window or use interpolation directly
    # Fill NAs in `stream_elevation` by using distance-based interpolation

    nearest_stream_elevation <- focal(stream_elevation, w = 9, fun = min, na.rm = TRUE)
     cat("Stream  focal elevation",  "\n")

    # Subtract interpolated stream elevation from DEM
    depth_to_water <- filledeelevation - nearest_stream_elevation

    # Mask the depth_to_water to get NA cells only
    na_mask <- is.na(depth_to_water)

    # Assuming depth_to_water is your raster with NA values
    non_na_mask <- !is.na(depth_to_water)  # TRUE for non-NA cells

    # Calculate the distance to the nearest non-NA cell
    distance_to_valid <- distance(non_na_mask)
   cat("Stream  distance",  "\n")  
    # Interpolate NA values based on nearby cells
    depth_to_water_filled <- focal(depth_to_water, w = 11, fun = mean, na.rm = TRUE)
     cat("dtw filled",  "\n")  
    # Smoothing using a mean filter
    r_smoothed <- focal(depth_to_water, w = matrix(1, 3, 3) / 9, fun = mean, na.rm = TRUE)


    # Replace Holes with Nearest Neighbor Values
    #Use terra::app to replace NA with the nearest non-NA value.

    r_filled_nn <- app(depth_to_water, fun = function(x) {
      ifelse(is.na(x), mean(x, na.rm = TRUE), x)
    })
    cat("app filled",  "\n")  

    # Crop the raster to the extent of the vector
    cropped_raster_dtw <- crop(r_filled_nn, huc8_boundary)
    # Mask the cropped raster to the vector's exact shape
    masked_raster_dtw <- mask(cropped_raster_dtw, huc8_boundary)
    cat("masked",  "\n")  
    # Combine DEM, slope, and aspect into one multi-layer raster
    combined_raster <- c(dem_cropped, slope,profile_curv,plan_curv,
                         sca,roughness,tpi,vrm,twi,
                     profile_curv9by9,profile_curv17by17,
                     #profile_curv100by100,
                     plan_curv9by9,plan_curv17by17,
                     #plan_curv49by49,
                     tpi_100m,tpi_200m,
                     #tpi_500m,
                     masked_raster_dtw)
    
     cat("combined",  "\n")  

    #  names(combined_raster) <- c( "elevation",  "slope"         ,    "profc"        ,     "planc"  ,
   #"flow_acc","twi", "roughness"      ,   "TPI"       ,        "vrm"  )
      # removing
   #  "profc_1000",
   #   "planc_490",
   #   "TPI_500" ,
    names(combined_raster) <- c( "elevation",  "slope"         ,    "profc"        ,     "planc"  ,
 "flow_acc", "roughness"      ,   "TPI"       ,        "vrm" , "twi",
 "profc_90","profc_170",
 
 "planc_90","planc_170",

 "TPI_100" ,"TPI_200" ,

 "dtw")
    huc8_name <- gsub("\\s+", "", huc8_name)
    # Save DEM as a GeoTIFF file
    file_name <- paste0(datapath,"/terr_stack", huc8_name, ".tif")
    writeRaster(combined_raster, file_name,  gdal=c("COMPRESS=LZW"),overwrite=TRUE)
    message(paste("DEM downloaded and saved for HUC8:", huc8_id," with ", huc8_name))
    gc()  # Free up memory
  }, error = function(e) {
    message(paste("Error downloading DEM for HUC8:", huc8_id, " - ", e))
    gc()  # Free up memory
  })
}

# Loop through all HUC8 regions in Washington and download DEMs
#for (i in 1:nrow(huc8_regions[1])) {
for (i in 2:nrow(huc8_regions[1])) {
  huc8_boundary <- huc8_regions[i, ]
  huc8_id <- huc8_regions$huc8[i]
   huc8_name <- huc8_regions$name[i]
  
  # Download DEM for each HUC8
  download_huc8_dem_name(huc8_boundary, huc8_id,huc8_name)
}
```


## Expected Outputs

Input DEM files in the /data/WA_10m directory. Multi-layer GeoTIFF files with terrain attributes. Logs of any errors encountered during processing.

## Tips and Troubleshooting

Ensure all packages are installed and up-to-date. Check file paths and permissions if data is not saving correctly. Review WhiteboxTools logs for specific tool errors.


